---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: true
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}

```

#### Load packages


##### lubridate

```{r, echo=TRUE, message=FALSE}
if(!"lubridate" %in% rownames(installed.packages())){
        install.packages("lubridate")}
if(!"lubridate" %in% rownames(.packages())){
        library(lubridate)}
```


##### dplyr

```{r, echo=TRUE, message=FALSE}
if(!"dplyr" %in% rownames(installed.packages())){
        install.packages("dplyr")}
if(!"dplyr" %in% rownames(.packages())){
        library(dplyr)}
```


##### ggplot2

```{r, echo=TRUE, message=FALSE}
if(!"ggplot2" %in% rownames(installed.packages())){
        install.packages("ggplot2")}
if(!"ggplot2" %in% rownames(.packages())){
        library(ggplot2)}
```


## Load and Preprocess Data

###1: Load CSV from [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)
```{r}
#create directory
if(!file.exists("./activitydata")){dir.create("./activitydata")}

#download data (zip)
if(!file.exists("./activitydata/activitydata.zip")){download.file("https://github.com/jcobb00/RepData_PeerAssessment1/raw/master/activity.zip",destfile = "./activitydata/activitydata.zip")}

#unzip data
if(!file.exists("./activitydata/activity.csv")){unzip("./activitydata/activitydata.zip",exdir = "./activitydata")}

#create activity data frame 
activity.data <- read.csv("./activitydata/activity.csv", header = TRUE)
```


###2: Preprocess the data

##### Format as date with lubridate

```{r}
activity.data$date <- ymd(activity.data$date)
```


##### Add day of week column

```{r}
activity.data <- cbind(activity.data,weekdays(activity.data$date))
colnames(activity.data)[4] <- "dayofweek"
```


##### Add weekday/weekend column

```{r}
activity.data <- cbind(activity.data,activity.data$dayofweek=='Saturday'|activity.data$dayofweek=='Sunday')
colnames(activity.data)[5] <- "weekend"
activity.data$weekend <- ifelse(activity.data$weekend == TRUE, "weekend", "weekday")
```


##### Use interval fraction to calculate ~time (interval/2355)

```{r}
activity.data <- cbind(activity.data,activity.data$date+(activity.data$interval/2355))
colnames(activity.data)[6] <- "date.time"
```


##### Reorder columns

```{r}
activity.data <- activity.data[,c(6,2,4,5,3,1)]
```


####Preprocessed data

```{r, message=FALSE}
head(activity.data)
str(activity.data)
```


## Daily Steps


###1: Calculate total steps taken each day

```{r, message=FALSE}
total.steps.per.day <- na.omit(activity.data) %>%
        group_by(date) %>%
        summarise(totalsteps=sum(steps))
print(total.steps.per.day)
```


###2: Make a histogram of the total number of steps taken each day

```{r total steps per day hist, message=FALSE, fig.width=8, fig.height=6}
ggplot(data=total.steps.per.day, aes(total.steps.per.day$totalsteps)) + 
        labs(title ="Total Steps per Day Histogram") +
        theme(plot.title = element_text(hjust = 0.5)) +
        xlab("Total Steps") +
        scale_y_continuous(breaks = seq(0, 20, by = 2)) +
        scale_x_continuous(breaks = seq(0, 25000, by = 5000)) +
        geom_histogram(binwidth = 2500)
```


###3: Calculate and report the mean and median total number of steps taken per day

```{r, message=FALSE}

daily.mean.median <- total.steps.per.day%>%
        summarise(mean=mean(totalsteps), median=median(totalsteps))
print(daily.mean.median)
```


## Average daily activity pattern

##### (first) Calculate average steps per interval across all days

```{r, message=FALSE}
interval.daily.avg <- na.omit(activity.data) %>%
        group_by(interval) %>%
        summarise(mean=mean(steps))
print(interval.daily.avg)
```


### 1: Plot 5-minute intervals and average steps taken across all days

```{r avg daily activity pattern, message=FALSE, fig.width=8, fig.height=6}
ggplot(data=interval.daily.avg, aes(x=interval, y=mean)) + 
        labs(title ="Average Daily Activity Pattern") +
        theme(plot.title = element_text(hjust = 0.5))+
        ylab("Steps") +
        xlab("Daily Interval") +
        scale_x_continuous(breaks = seq(from=0,to=2400,by=300)) +
        geom_line(stat="identity")
```


### 2: Largest 5 minute interval averaged across all days

```{r}
arrange(interval.daily.avg,desc(mean))[1,]
```


## Imputing missing values


### 1: Calculate and report the total number of missing values in the dataset

```{r}
as.numeric(sum(is.na(activity.data$steps)))
```

### 2: Devise a replacement plan
        All NAs occur on 8 days, use average interval for those days

```{r}
daily.nas <- activity.data %>%
        group_by(date) %>%
        tally(is.na(steps)) %>%
        arrange(desc(n))
daily.nas[1:8,]
```


### 3: Set NAs to interval average

```{r}
mean.na <- activity.data
mean.na[is.na(activity.data)] <- mean(activity.data$steps,na.rm = TRUE)
head(mean.na)
```


### 4(a): Plot a histogram of total steps per day (na=mean)

```{r avg steps per day hist na mean, fig.width=8, fig.height=6}
daily.mean.median.na <- mean.na %>%
        group_by(date)%>% 
        summarise(mean=mean(steps), median=median(steps))

ggplot(data=daily.mean.median.na, aes(daily.mean.median.na$mean)) + 
        labs(title ="Avg Steps per Day Histogram (na=mean)") +
        theme(plot.title = element_text(hjust = 0.5)) +
        xlab("Avg Steps") +
        stat_bin(binwidth = 10)
        geom_histogram()
```


### 4(b): Calculate daily mean and median (na=mean)
        Mean does not change, median becomes the mean

```{r}
daily.mean.median.na <- mean.na %>%
        group_by(date) %>%
        summarise(totalsteps=sum(steps)) %>%
        summarise(mean=mean(totalsteps), median=median(totalsteps))
print(daily.mean.median.na)
```


## Differences in activity patterns between weekdays and weekends

### 1: Create a new factor variable: "weekday" and "weekend" (created during preprocessing)

```{r}
if(!"weekend" %in% colnames(activity.data)) {
        activity.data <- cbind(activity.data,activity.data$dayofweek=='Saturday'|activity.data$dayofweek=='Sunday')
        colnames(activity.data)[5] <- "weekend"
        activity.data$weekend <- ifelse(activity.data$weekend == TRUE, "weekend", "weekday")}
```


### 2: Panel plot of interval steps averaged across all days and grouped by weekday and weekend

##### (first) Calculate average interval steps across weekdays and weekends

```{r}
avg.day.of.week <- na.omit(activity.data) %>%
        group_by(interval,weekend) %>%
        summarise(avgsteps=mean(steps)) %>%
        arrange(desc(weekend))
```

```{r avg interval per weekday weekend, fig.width=8, fig.height=6}
ggplot(data=avg.day.of.week, aes(x=avg.day.of.week$interval, y=avg.day.of.week$avgsteps)) + 
        labs(title ="Average Interval per Weekday/Weekend") +
        theme(plot.title = element_text(hjust = 0.5)) +
        facet_grid(weekend ~ .) +
        ylab("Steps") +
        xlab("Daily Interval") +
        geom_line(stat="identity")
```


